Substitution rate responses to changes in effective population size.

### 2020-04-22 manuscript/manuscript.tex
Refactored manuscript

Differences in DNA sequences of a gene in separate species are due to the particular history of DNA substitutions along each specie's respective lineage.
These substitutions at the level of the population are the result of point mutations at the level of individuals, subject to evolutionary forces such as selection and random drift.
In protein-coding DNA sequences, non-synonymous mutations are changing the amino-acid sequence, and by comparing the non-synonymous substitutions rate relative to the synonymous substitution rate (supposedly neutral), a statistic called $\dnds$, we can estimate the global strength of selection and random drift.\\
For a given gene, it has been empirically observed changes of $\dnds$ along lineages of a phylogeny \cite{Yang2001, Zhang2004}.
More importantly, changes in substitution rate correlate with changes in life-history-traits (longevity, body mass, ...) such as shown by the molecular comparative framework \cite{Lartillot2011,Weber2014}.
Such covariations are compatible with the theoretical prediction of the nearly neutral theory of evolution \cite{Ohta1972, Ohta1992}, where the underlying mechanism is changes in effective population size ($\Ne$).
Indeed, populations with low $\Ne$ would have both a large body size and a high $\dnds$ (weaker selection) due to the increase of random drift.
If $\dnds$ is empirically correlated to life-history traits, such as body mass and longevity, the direction of correlation could not be replicated across all experiments \cite{Figuet2016}.\\
Theoretical prediction of $\dnds$ decreasing as a function of $\Ne$ had been developed in the context of a fixed distribution of fitness effects of novel mutations \cite{Ohta1972, Welch2008}. Similarly, it has been shown in context of a fixed fitness landscape, where each amino-acid have different fitness \cite{Spielman2015a, DosReis2015}, although sites are considered independently evolving, without epistasis.
Unexpectedly, it has also been argued that the existence of a $\dnds$ response to $\Ne$ is not straightforward, even in the context of nearly neutral theory of evolution \cite{Lanfear2014}.
In the example of a biochemical mechanistic computational model, where the fitness is determined by the probability of folding, $\dnds$ as been observed to be independent of $\Ne$ \cite{Goldstein2013}.
Theoretically, a more general case is obtained whenever the fitness is a concave function of a trait, and such trait is equimutable \cite{Cherry1998}.
Equimutabibilty meaning a mutation has the same effect on the trait whatever the current trait of the individual.
In addition to this complexity, $\dnds$ response to $\Ne$ can be non-monotonic due to non-equilibrium properties, such that an increase in $\Ne$ implies firstly a burst of $\dnds$ (adaptive selection), and subsequently a decrease in $\dnds$ (purifying selection) \cite{Jones2016}.\\
Our goal in this present work to reconcile empirical observation with theoretical results, such as to delimit which models and assumptions are reasonable.
This question is important if we want to extract changes in effective population size from substitution history, as well as to determine the fitness landscape under which protein-coding DNA sequences are evolving.
We provide theoretical tools to derive the relationship between $\Ne$ and $\dnds$ in the context of a genotype-phenotype-fitness map.
\section*{Results}
\subsection*{Theoretical approximations}
In this section we theoretically approximate the change in $\dnds$ at equilibrium after a change in $\Ne$, under a model in which the fitness depends on the genotype of the whole sequence.
Firstly, we decompose the map from genotype to fitness through an intermediary bounded phenotype $0 \leq \x \leq 1$, the proportion of unstable sites in the sequence.
Each site of the sequence ($\Nsite$ sites) can be in one of $\Nstate \geq 2$ states, where only $1$ of those states is assumed to be a stable state, and $\Nstate - 1$ states are unstable.
The optimal phenotype is obtained when all of the $\Nsite$ sites of the sequence are stable ($\x = 0$).\\
To complete the model, the phenotype-fitness map is first considered generally, where the Wrightian fitness of phenotype $\x$ is $f(\x)$.
After a mutation, given that only one site can change at a time, the absolute change of $\x$ is either $0$ or $\dx=\sfrac{1}{\Nsite}$.
Most importantly the probabilities to increase or decrease $\x$ are different, and depends on the current phenotype $\x$.
Mutation increasing $\x$ are selected against, while mutations decreasing $\x$ are positively selected. 
Mutation and selection cancel each other at equilibrium, where the expectation of fitness effects of mutations reaching fixation in the population is $0$.
Altogether, the equilibrium phenotype obtained for such mutation-selection balance is denoted $\x\eq$, and more importantly is given by the equation (See Supplementary Materials):
\begin{align}
\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}.
\end{align}
This equation of equilibrium phenotype can be further studied to derive the response of $\x\eq$ after a change in $\Ne$, using differential calculus.
Furthemore, assuming that the number of states is large (typically $20$ for amino-acid), we obtain the change in $\omega\eq$ after a change in $\Ne$ as: 
\begin{align}
\frac{ \der \omega\eq}{\der \ln (\Ne)} & \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}.
\end{align}
In other words, the slope of the $\dnds$ elasticity to changes in $\Ne$ (in log scale) is the ratio of first to the second derivative of the log-fitness function, taken at the equilibrium phenotype. To note, this elasticity is strictly negative for decreasing log-concave fitness functions.\\
So far, our result did not assume any particular parametrization of the phenotype-fitness map.
As an example, already proposed in the literature \cite{Goldstein2011}, fitness is proportional to the probability of our protein to be in the folded state, given by Fermi Dirac distribution: 
\begin{equation}
f(\x) \propto \dfrac{1}{1 + e^{\beta (\deltaGMin + \Nsite \gamma \x)}}, 
\end{equation}
where $\deltaGMin < 0$ and $\gamma > 0$ are in kcal/mol, and the fixed parameter $\beta$ is $1.686$ mol/kcal at room temperature. In other words, $\deltaGMin$ is the difference in free energy between folded and unfolded state when all sites are stable. $\gamma$ is the expected change in free energy (between folded and unfolded states) for a destabilizing mutation.
With such fitness function, the equilibrium phenotype $\x\eq$ is given by the equation: 
\begin{align}
\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq 4\Ne \beta \gamma e^{\beta (\deltaGMin + \Nsite \gamma \x\eq)}.
\label{eq:equilibrium}
\end{align}
The $\dnds$ elasticity to changes in $\Ne$ simplify to a compact equation: 
\begin{equation}
\frac{ \der \omega\eq}{\der \ln (\Ne)} \simeq -\dfrac{1}{\beta \Nsite \gamma};
\end{equation}
which is independent of $\x\eq$, meaning $\omega$ is linearly decreasing with $\Ne$ in log space.
Moreover, only the free compound parameter $\gamma \Nsite$ has an impact on the slope, an intuition for such result is given in figure \ref{fig:NeChangeInfluence}.
Meaning that under empirically relevant value of $\Nsite=300$ sites and $\gamma=1.0$ kcal/mol \cite{Zeldovich2007}, the slope of the linear relationship between $\dnds$ and log-$\Ne$ is approximately $0.002$.
  \includegraphics[width=0.9\textwidth, page=1] {artworks/theoretical.pdf}
   \textbf{Theoretical model.}
   Panel A. Illustration of the Genotype-phenotype-fitness map. The phenotype ($\x$) is a real-valued summary of the genotype, and is defined in our model as the fraction of unstable sites in the sequence. The fitness is a decreasing log-concave function of the phenotype.
   Panel B. Illustration of the $\dnds$ elasticity after a change in $\Ne$.
   For the special case $K=2$, the equilibrium $\x\eq$ is determined by the equation $\ln(\frac{1 - \x\eq}{\x\eq})=4\Ne \beta \gamma e^{\beta (\deltaGMin + \Nsite \gamma \x\eq)}$. The right-hand side of the equation increases exponentially with $\x$ where $\Nsite \gamma$ is the exponential growth rate. Thus when $\Nsite \gamma$ is large (red solid line), increasing $\Ne$ (red dotted line) only moves slightly $\x\eq$ (x-axis of the crossing with the blue solid line). On the other hand, when $\Nsite \gamma$ is low (green solid line), increasing $\Ne$ (green dotted line) moves $\x\eq$ notably. Moreover, changes in $\x\eq$ reflects the changes in $\dnds\eq$ since both are related by a monotonous function (See Supplementary Materials). The value of $\alpha$ has been chosen given all other parameters such that the solid lines all cross at the same point.
  \label{fig:NeChangeInfluence}

\subsection*{Simulations confirmation}
Our compact theoretical results are based on assumptions that are broken in practice, for example assuming that 
\begin{enumerate}
 \item $\Nsite$ is large;
 \item selection coefficient is well approximated by the fitness derivative;
 \item genetic code is not taken into account;
 \item all unstable states are equivalently unstable.
\end{enumerate}
We thus first tested the soundness of our theoretical results when such assumptions are relaxed, by simulating DNA sequences evolution (See Methods). 
Firstly, simulations demonstrate that the relation between $\dnds$ and log-$\Ne$ is linear, and that the slope of the linear regression match the expected theoretical value (See Figure \ref{fig:GoldsteinVsToy}).
Secondly the parameter $\deltaGMin$ hardly changes the slope of the linear regression, as expected also theoretically.
Decreasing $\deltaGMin$ (to more negative values) increases $\dnds$, by shifting the equilibrium to higher $\x\eq$ since more unstable sites are fixed before reaching sensible deleterious selection coefficient against unstable mutations. Once many sites are unstable, the $\dnds$ is higher since non-synonymous mutations between unstable states are effectively neutral. However the slope of the $\dnds$-$\Ne$ relationship is not changed, a predicted in our theoretical model.\\

Finally, we compare our results to more complex models of sequence evolution leveraging the $3$D structure of protein \cite{Williams2006, Goldstein2011, Pollock2012}.
In such models, the free energy of the folded state is computed using the $3$D folded structure and pairwise contact potential energies between neighboring amino-acid residues \cite{Miyazawa1985}.
The free energy distribution of unfolded states is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states (see Supplementary Materials).
The original works claimed that under such $3$D model $\dnds$ is independent of $\Ne$.

@@

In such experiment, $\alpha=-118$ kcal/mol as it is the $\deltaG$ of the most stable sequence of $300$ sites \cite{Goldstein2011}.

@@

\subsection*{Epistasis determines the time to relaxation}

@@

\subsection*{Distribution of fitness effects}
DNA mutations changing a genotype can result in a change of phenotype, and ultimately a change in fitness.
From a specific genotype, all the possible mutations thus result in a distribution a phenotypic effect (DPE) and fitness effects (DFE).
The DPE and DFE are not known a priori, but are the resulting consequence of the mutation-selection-drift balance (see Figure \ref{fig:Distribution}).
Empirically, these distributions are of particular importance since they can be obtained experimentally or inferred with other data.
As an example, DFE can be inferred from polymorphism dataset \cite{Eyre-walker2007, Galtier2016}. 
Moreover, the distribution of $\deltadeltaG=\gamma \Nsite \dx$ of novel mutations can be obtained experimentally.
  \includegraphics[width=0.9\textwidth] {artworks/DPE-DFE.pdf}
   \textbf{Distribution of phenotypic and fitness effects.}
   Distribution of fitness effects and phenotypic effect for novels non-synonymous mutations observed along a simulation at the mutation-selection balance. $\alpha=-118$, $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
   Each side of the distribution is fitted to a gamma distribution, shown in dotted line. 
   Panel A. Distribution of observed $\deltadeltaG$, which fit adequately the gamma distribution for negative $\deltadeltaG$ (stabilizing mutations). 
   Panel A. Distribution of observed selection coefficient, which fit adequately the gamma distribution for both positive and negative selection coefficient. However the shape parameter estimated is not the same for positive and negative selection coefficients. 
  \label{fig:Distribution}
\section*{Discussion}
Goldstein \& Pollock argued that model of protein evolution should consider how the acceptance of mutations depends on the sequence context in which they arise (epistasis), as to correctly represent resultant rate heterogeneity along the sequence, to predict the role of compensatory substitutions in protein evolution, predict which of the 10\% of deleterious mutations in humans are harmless in other species, or to accurately represent the rate and time dependence of convergence and homoplasy \cite{Goldstein2017}.
We argue that not taking into account epistasis also have the consequence of overestimating the relaxation time to return to equilibrium of $\dnds$ after changes in $\Ne$.
Paradoxically, not taking into account epistasis means overestimating the response of $\dnds$ to changes in $\Ne$ at equilibrium.\\
If using a distribution of fitness effect (DFE) is mathematically convenient and does not suffer from underestimating the relaxation rate, it is also overestimating the response of $\dnds$ to changes in $\Ne$ at equilibrium.
As such, developing inference of fitness landscape will require to assume site-interdependent fitness landscapes.\\
\section*{Materials \& Methods}
Protein sequence evolution is simulated under an origin-fixation model \cite{McCandlish2014}., where the rate of substitution of a new mutation with selection coefficient $s$ is:
\begin{equation}
p_{\text{fix}}(s) = \mu \dfrac{4 \Ne s}{1 + e^{- 4 \Ne s }}, 
\end{equation}
where $\mu$ is the mutation rate.
\subsection*{Simulation of protein folding with an additive model of free energy}
\label{MatMet:folding}
From a DNA sequence $\ci$ after $t$ substitutions, the protein's difference in free energy between folded and unfolded state is given by:
\begin{equation*}
\deltaG\left(\ci\right) = \deltaGMin + \Nsite \gamma * \x\left(\ci\right), 
\end{equation*}
where $\x\left(\ci\right)$ is the proportion of non-optimal amino-acid sites in the sequence.
Wrightian fitness is defined as the probability of our protein to be in the folded state, given by the Boltzmann equation: 
\begin{equation}

>>>>

Molecular sequences differ across species due to the particular history of DNA substitutions along each specie's respective lineage.
These substitutions in molecular sequences are the result of the interplay between evolutionary forces such as mutation, selection and random genetic drift.
These forces have effects at different levels, mutations are carried by molecular sequence, selection is beared by individual's fitness, and random genetic drift is a population effect.
The strength of random genetic genetic, or stochastic sampling of mutations, is less pronounced in lineage with high effective population size ($\Ne$), and as a consequence the purification by selection of deleterious mutations is more effective.
Such idea has emerged in the framework of quasi-neutral theory of evolution, whereby predicting a decrease in the substitution rate per unit of time of selected mutations along a lineage with higher $\Ne$ \cite{Ohta1972, Ohta1992}.
This prediction is valid under the assumption that most mutations are deleterious and that their selective effect on individuals are drawn from a fixed distribution \cite{Welch2008}.
Furthermore, this prediction assumes a constant mutation rate per unit of time, but such assumption can be relaxed by normalizing the substitution rate of selected mutations by the substitution rate of neutral mutations such as to discard this confounding factor \cite{Ohta1972}.
This relative substitution rate of selected mutations, called $\dnds$ is this study, is a theoretical macroscopic observable in molecular sequence evolution, given we can separate selected and neutral mutations.
Practically, $\dnds$ can be approximated in protein-coding DNA sequences, where selected mutations are the non-synonymous changes affecting the amino-acid sequence, and the non-synonymous mutations are supposedly neutral, this statistic is called $dNdS$.\\
% \dnds decrease with \Ne under the assumption of fixed DFE

The context of protein coding sequence fostered another modeling of selection, based on genotype-fitness map instead of distribution of fitness effects.
In such approach, the selective effect of a mutation depends on the fitnesses of both source and target amino-acids involved on the mutation \cite{Halpern1998}.
For example, if the target amino-acid has an higher fitness, the selective effect of the mutation is positive, and reciprocally negative for lower fitness.
Such modeling approach account for higher substitutions rate between amino-acid with similar bio-chemical properties.
Fundamentally, in such framework the fitness depends solely on the actual genotype, not on the trajectory of mutations leading to this sequence.
Even thought the modeling approach is substantially different, assuming each site of the sequence has its own genotype-fitness map also predicts a negative correlation between $\dnds$ and $\Ne$ \cite{Spielman2015a, DosReis2015}.\\
% \dnds decrease with \Ne under the assumption site fitness landscape

Empirically, fluctuation of $\dnds$ along lineages of phylogenies have been inferred using protein-coding DNA alignments \cite{Yang2001, Zhang2004}.
Subsequently, inference methods combining molecular sequences and lineage specific quantitative traits have asserted that $\dnds$ correlates positively with longevity and body mass, traits which have been sampled only in the extant lineages \cite{Lartillot2011,Weber2014}.
Moreover, longevity and body mass are proxies of $\Ne$, with low $\Ne$ corresponding to lineage with a large body size and extended longevity \cite{Romiguier2014}.
As such, empirical evidences suggest a negative correlation between $\dnds$ and $\Ne$, confirming the theoretical prediction of the quasi-neutral theory of evolution.
However, this correlation could not be replicated across all experiments, and depending on the clade or the potential biases taken into account, the correlation could be in the inverse direction or not statistically significant \cite{Figuet2016}.\\
% \dnds decrease with \Ne empirically, but not always

Mitigated empirical evidence that $\dnds$ decrease with $\Ne$ encouraged alternative explanatory mechanism \cite{Lanfear2014}.
One striking result proved that $\dnds$ can independent of $\Ne$, when fitness is a function of a phenotype and that such phenotype is equimutable.
Equimutability states that the probabilities of phenotype changes due to mutations is independent the current phenotype of the individual \cite{Cherry1998}.
Such argument has been invoked in the \textit{in silico} observation that $\dnds$ is seamingly independent of $\Ne$ if the protein thermodynamic stability is computed using a 3D structural model of the protein, and that the fitness is proportional to the folding probability of the protein \cite{Goldstein2013}.
The given explanation is that probabilities of changes in free energy of folding ($\deltadeltaG$) is independent of the current free energy ($\deltaG$), which is considered an equimutable phenotype.\\
% Explanation could be that \dnds is independent of \Ne

However, this assumption is relatively strong, as for example if a protein is completely stable, only destabilizing mutations can occur, and more generally the probabilities of destabilizing mutational events is dependent on the actual composition of the protein.
Under the assumption that fitness relates to protein stability, and that such stability if a function of the genotype, the theoretical $\dnds$ relationship to $\Ne$ is unknown.
The goal of this study is to derive the $\dnds$ relationship to $\Ne$ to intrinsic molecular parameters of the model.
More generally, we develop mathematical tools to derive the $\dnds$ relationship to $\Ne$ in the context of a genotype-phenotype-fitness map.
With such results, we stress to test the assumption that protein stability is a proxy of fitness in the light of empirical data.\\
% The assumptions are strong, we want to relax equimutability which is unlikely

Orthogonally to the relationship between $\dnds$ and $\Ne$, independent studies have argued that protein abundance is also a predictor of $\dnds$.
As such, we also seek to derive the $\dnds$ relationship to protein abundance, and discuss the empirical implication.\\
% \dnds is relationship to protein abundance

Lastly, the theoretical results discussed so far are all obtained when the balance between mutation, selection and drift is at equilibrium.
However, under a model of site-dependent genotype-fitness map, an increase in $\Ne$ leads first to an increase of $\dnds$ due to adaptive selection, and subsequently a decrease in $\dnds$ due to stronger purifying selection \cite{Jones2016}.
Studying only equilibrium properties can thus be misleading, and the dynamical response of $\dnds$ to changes in $\Ne$ must also be addressed by our model.\\
% Result at equilibrium, what about dynamical properties?

\section*{Results}
\subsection*{Theoretical approximations}
In this section we present an analytical approximate solution for the change in $\dnds$ at equilibrium after a change in $\Ne$.
The main results are given in a general case of any phenotype-fitness map, and also derived in the specific case of fitness proportional to the total amount of folded protein.
All development are available in Supplementary Materials in the most general case.
The first step define the relationship from genotype to fitness, through an intermediary phenotype.
The second step present the equation of the phenotype at equilibrium, when evolutionnary forces of mutation, selection and genetic drift compensate each others.
Finally, an approximate response of $\dnds$ to changes in $\Ne$ is derived from the equilibrium equation. \\

We base our work on numerical model of protein stability \cite{Williams2006, Goldstein2011, Pollock2012}, on which we make simplifying assumptions to derive analytical equations.
In the original model, the protein stability is the difference in free energy between folded and unfolded conformations, called $\deltaG$ in kcal/mol.
Technically, the free energy of the folded are unfolded conformations are computed using $3$D conformations of the protein.
In such model, the stabilizing or destabilizing effect of an amino-acid at a particular site depends on the amino-acids present in the vicinity in $3$D conformation.\\

We approximate this model such that the destabilizing effect of an amino-acid does not depend on a $3$D representation of the protein, but each site contribute independently to $\deltaG$.
With such approximation, at a given site of sequence each amino-acid is either stabilizing or destabilizing the protein, and an unstable amino-acid contribute to an excess of $\gamma > 0$ (in kcal/mol) to $\deltaG$ compare to a stable amino-acid.
If all amino-acids of the sequence are stable, $\deltaG$ equal to $ \alpha < 0$. 
In this model, the more compact phenotype summarizing the genotype is the proportion of unstable amino-acid in the sequence, defined as $0 \leq \x \leq 1$, and $\deltaG$ is a linear function of $\x$:
\begin{align}
\deltaG (\x) = \alpha + \Nsite \gamma \x,
\end{align}
where $\Nsite$ is the number of sites in the sequence. \\

From a given genotype, mutations changing the sequence have various effect, they can increase or decrease the proportion of unstable amino-acid, or do nothing if the mutation is between both unstable (or equivalently stable) amino-acids.
To derive the probabilities of such events to occur, we also make the simplyfing assumption that the transition between amino-acids are equiprobable and only one amino-acid is stable and the $19$ others are unstable.
Altogether, any mutation in the sequence can then have a phenotypic effect of $0$ or $\dx=\sfrac{1}{\Nsite}$, and the probabilities of transition are:
\begin{gather}
\begin{cases}
\dx &\text{ with probability } 1-x, \\
0 &\text{ with probability } \frac{18 \x }{19}, \\
-\dx &\text{ with probability } \frac{\x}{19}.\\
\end{cases} \label{eq:proba}
\end{gather}
In the extreme case of optimal phenotype ($\x = 0$), only destabilization mutations are proposed.
Moreover, the probability to propose stabilizing mutation (effect $-\dx$), or between unstable amino-acids (effect $0$) is proportional to $\x$. 
Once obtained the phenotypic effects of mutations and their probabilities, definition of the phenotype-fitness map is necessary derive the selective effect of mutations.\\

Subsequent equations are valid for the more general case where the Wrightian fitness is a log-concave differentiable function $f$ of phenotype $\x$.
As an example, we develop the argument when the Wrightian fitness of an individual is the proportion of protein in folded conformation.
This fitness function is motivated in part that a protein must be folded to perform its function, but can also be justified by the toxic effect of misfolded proteins in the cytoplasm. 
This fitness is then given by the Fermi Dirac distribution and is typically close to $1$, leading to a first-order approximation\cite{Goldstein2011}: 
\begin{align}
f(\x) = \dfrac{1}{1 + e^{\beta (\alpha + \Nsite \gamma \x)}},\\
\Rightarrow f(\x) \simeq 1 - e^{\beta (\alpha + \Nsite \gamma \x)}, 
\end{align}
where $\alpha$ and $\gamma$ are defined as above, and the fixed parameter $\beta$ is $1.686$ mol/kcal at room temperature.
Destabilizing mutations are selected against with a negative selection coefficient which can be approximated by:
\begin{align}
s & \simeq \frac{1}{\Nsite}\frac{ \partial \ln f(\x) }{\partial \x} \label{eq:s} \\
\Rightarrow s & \simeq - \beta \gamma e^{\beta (\alpha + \Nsite \gamma \x)} \label{eq:s-unfolded}
\end{align}

With the complete genotype-phenotype-fitness map defined (see Figure \ref{fig:NeChangeInfluence}, left panel), one can study the evolutionary processes unfolds.
Starting from an optimal sequence, destabilizing mutations will reach fixation and accumulate until the selection coefficient against new deleterious mutations is too strong, a point of equilibrium called marginal stability.
Most importantly the probability of fixation of mutations is affected by genetic drift, parameterized by effective population size ($\Ne$).
Altogether, at the equilibrium between mutation, selection and drift, the selection coefficient of new advantageous and deleterious mutations reaching fixations are expected to be null on average.
Formally, and after simplification, the equilibrium phenotype denoted $\x\eq$ is given by the equation:
\begin{align}
\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (19) & \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}, \\
\Rightarrow \ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (19)  & \simeq 4\Ne \beta \gamma e^{\beta (\alpha + \Nsite \gamma \x\eq)}. \label{eq:equilibrium}
\end{align}
This equation cannot be solved explicitly for $\x\eq$, but an intuition on the consequences of change in $\Ne$ to the equilibirum phenotype $\x\eq$ is given in Figure \ref{fig:NeChangeInfluence} (right panel). \\
\begin{figure*}[htb!]
	\begin{mdframed}
		\centering
		\includegraphics[width=0.9\textwidth, page=1] {artworks/theoretical.pdf}
		\caption{
			\textbf{Theoretical model.}
			Panel A. Illustration of the Genotype-phenotype-fitness map. The phenotype ($\x$) is a real-valued summary of the genotype, and is defined in our model as the fraction of unstable sites in the sequence. The fitness is a decreasing log-concave function of the phenotype.
			Panel B. Illustration of the $\dnds$ elasticity after a change in $\Ne$. The equilibrium $\x\eq$ is determined by the equation $\ln(\frac{1 - \x\eq}{\x\eq}) + \ln(19)=4\Ne \beta \gamma e^{\beta (\alpha + \Nsite \gamma \x\eq)}$. The right-hand side of the equation increases exponentially with $\x$ where $\Nsite \gamma$ is the exponential growth rate. Thus when $\Nsite \gamma$ is large (red solid line), increasing $\Ne$ (red dotted line) only changes subtly $\x\eq$ (x-axis of the crossing with the blue solid line). On the other hand, when $\Nsite \gamma$ is low (green solid line), increasing $\Ne$ (green dotted line) involve a stronger response of $\x\eq$. Moreover, changes in $\x\eq$ reflects the changes in $\dnds$ since both are related by a monotonous function (equation \ref{eq:dnds}). The value of $\alpha$ has been chosen given all other parameters such that the solid lines all cross at the same point.
		}
		\label{fig:NeChangeInfluence}
	\end{mdframed}
\end{figure*}

Moreover, at equilibrium we can derive the expected substitution rate of mutations changing the sequence, and by normalizying by the substitution rate of neutral mutations, $\dnds$ simply approximates to:
\begin{gather}
\dnds \simeq \x\eq \label{eq:dnds}
\end{gather}
This simple approximation is due to substitutions at unstable sites (proportion $\x\eq$) to one of the other unstable amino-acids, which are effectively neutral and compose the largest proportion of proposed mutations with substantial probability of fixation (see equation \ref{eq:proba}).\\

From the equilibrium equation \ref{eq:equilibrium}, differential calculus allow to derive the response of $\x\eq$ and ultimately $\dnds$ after a change in $\Ne$.
We define the elasticity $\chi$ as slope of the $\dnds$ response to changes in $\Ne$ in log scale, which simplify to a compact equation: 
\begin{align}
\chi = \frac{ \der \dnds}{\der \ln (\Ne)} & \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}, \\
\Rightarrow \chi & \simeq -\dfrac{1}{\beta \Nsite \gamma}.
\end{align}
The elasticity is the ratio of first to the second derivative of the log-fitness function, taken at the equilibrium phenotype. To note, this elasticity is strictly negative for decreasing log-concave fitness functions, asserting that $\dnds$ is a decreasing function of $\Ne$.
In the case of fitness equal to the folded proportion, $\chi$ is independent of $\x\eq$, meaning that $\dnds$ is linearly decreasing with $\Ne$ in log space.
Moreover, only the compound parameter $\beta \gamma \Nsite$ has an impact on the slope of the linear relationship, where only $\gamma$ and $\Nsite$ are free parameters. \\

Other parametrization of the fitness function shed light on the robustness of the theoretical results.
If the selective coefficient itself is not equal (equation \ref{eq:s-unfolded}) but proportional to the total amount of misfolded protein, and the expression level of the protein is denoted $y$, then the response of $\dnds$ to changes in $\Ne$ is the same as the response to changes in $y$ (See Supplementary Materials):
\begin{align}
s & \propto - \beta \gamma y  e^{\beta(\alpha + \gamma \Nsite \x)} \\
\Rightarrow \chi = \frac{ \der \dnds}{\der \ln (y)} & \simeq -\dfrac{1}{\beta \Nsite \gamma}.
\end{align}
Meaning we should observe the same relationship between $\dnds$ and $\Ne$ that between $\dnds$ and expression level.\\

Under empirically relevant value of $\Nsite=300$ sites and $\gamma=1.0$ kcal/mol \cite{Zeldovich2007}, the elasticity $\chi \simeq -0.002$.
In other words, for a increase in $\Ne$ of $4$ orders of magnitude, $\dnds$ decrease approximately of $0.01$, a subtle relationship that requires laborious  effort to be detected in simulated data, and large amount of empirical data with few bias and noise to be detectable.

\subsection*{Simulations confirmation}
In order to test the soundness of our compact theoretical results, we relaxed several assumptions by simulating DNA sequences evolution (See Methods), under various range of $\Ne$ and reporting the average $\dnds$ observed along the computation. 
With respect to mutation, we assumed that all amino-acids transition are equiprobable, in other word the complexity of the genetic code is not taken into account.
Simulating evolution of DNA sequence, and taking into account a matrix a mutation rate between nucleotide allow to test directly the robustness of the results to this assumption.
Furthermore, with regards to fitness, we assumed that all unstable amino-acids are equivalently unstable.
This assumption is relaxed by choosing for every site of the sequence one stable amino-acid, and the impact on the phenotype ($\x$) of other amino-acids is not binary, but scaled by the Grantham amino-acid distance \cite{Grantham1974}.
Also, the number of sites in the sequence ($\Nsite$) is assumed to be large such that the selection coefficient is well approximated by the fitness derivative (equation \ref{eq:s}).
The robustness of such approximation is test in the simulations with sequences of a finite number of sites ($\Nsite=300$).\\

Simulations demonstrate that the relation between $\dnds$ and log-$\Ne$ is linear, and that the slope of the linear regression match the expected theoretical value (See Figure \ref{fig:GoldsteinVsToy}, panel B).
Secondly we observe that effect of the parameter $\alpha$ is subtle on the slope of the linear regression, as expected also theoretically (See Figure \ref{fig:GoldsteinVsToy}, panel A).
Decreasing $\alpha$ (to more negative values) increases $\dnds$, by shifting the equilibrium to higher $\x\eq$ since more unstable sites are fixed before reaching the point of marginal stability.\\

Finally, we relaxed our assumption that each site of the sequence contribute independently to $\deltaG$, by taking into account the $3$D structure of protein.
We implemented the original model \cite{Williams2006, Goldstein2011, Pollock2012}, in which the free energy of the folded are unfolded conformations are computed using $3$D structures of the protein conformations and pairwise contact potential energies between neighboring amino-acid residues \cite{Miyazawa1985} (see Supplementary Materials).
The original works claimed that under such $3$D model $\dnds$ is independent of $\Ne$ \cite{Goldstein2013}.

@@

In such experiment, $\alpha=-118$ kcal/mol as it is the $\deltaG$ of the optimal sequence of $300$ sites \cite{Goldstein2011}.

@@

\subsection*{Epistasis and time to relaxation}

@@

Previous studies have argued that model of protein evolution should consider epistasis for various empirical reasons such as rate heterogeneity along the sequence, rate and time dependence of convergence, role of compensatory substitutions, prediction of deleterious mutations \cite{Goldstein2017}.
We argue that epistasis also have an important role the elasticity of $\dnds$ to changes in $\Ne$.
If epistasis is not taken into account, the elasticity of $\dnds$ is spread along time incompatible with biological evolution. \\
% Epistasis influence relaxation time

\section*{Discussion}
The negative relationship between the relative substitution rate of selected mutations ($\dnds$) and $\Ne$ has been predicted theoretically under the quasi-neutral theory of evolution.
Empirical data showed mitigated confirmation of this prediction, encouraging alternative explanatory mechanism.
One such explanation, developed in the context of genotype-phenotype-fitness map, demonstrated the absence of $\dnds$ elasticity to changes in $\Ne$ if the probabilities of phenotype changes are invariant to the current phenotype.
We argue that this assumption can be relaxed, and provide theoretical tools to derive the relationship between $\Ne$ and $\dnds$ in this context.
We apply our framework in the special case of fitness proportional to the proportion of folded proteins. \\
% Goal and theoretical results

Our theoretical results demonstrate that the relationship between $\dnds$ and $\Ne$ (in log space) is linear with a negative slope.
This slope is a function of molecular parameters of the model, and is inversely proportional to the product of the sequence size and the average change in conformational energy of destabilizing mutations. 
Based on empirical estimate of the molecular parameters, the elasticity of $\dnds$ to changes in $\Ne$ would be $\chi \simeq 0.002$.
Our compact theoretical results are supported by more complex simulations of protein evolution relaxing several assumptions.
% Our theoretical results are robsut to approximations

Particularly, our theoretical prediction match a numerical model of protein evolution, in which the free energy of the folded are unfolded conformations are computed using $3$D structures of the protein conformations.
Previous studies using this model presented an apparent lack of $\dnds$ elasticity to changes in $\Ne$ \cite{Goldstein2013}.
We argue the apparent lack of elasticity is in fact due to the subtle relation, and require extensive computation to be detected.\\
% Goldstein \& Pollock model don't have a null elasticity

Empirically, inference of the $\dnds$ elasticity in Primates, using polymorphism and divergence data estimated $\chi \simeq $, approximately $X$ times greater than our theoretical prediction \cite{Brevet2019}.
More empirical data across clades would be required to infirm such model, but in the light of empirical data estimating the relationship between $\Ne$ and $\dnds$, models based on the probability of folding are not reasonably fitted.\\
% Our result show the fitness proportional to proportion of folded folding is at odd with empirical data

Furthermore, our theoretical results manifest that the relationship between $\Ne$ and $\dnds$ is the same as the relationship between protein abundance and $\Ne$, whenever protein fitness is determined by the deleterious effect of unfolded proteins.
As such, dataset relating $\dnds$ to $\Ne$ and protein can be used confirm or infirm this model.\\
% Stress the importance that elasticity is valid for Ne and protein abondance

Notably, theoretical approximations apply more broadly to protein-protein interactions (See Supplementary Materials), where protein may either be in free form or engaged in a non-specific interaction. 
In non-specific interactions, stable site are hydrophilic amino-acid and unstable site are hydrophobic amino-acid at the protein surface.
Fitting this model with empirical estimates \cite{Zhang2008}, we obtain an elasticity of $\chi = 0.2$ (See Supplementary Materials) thus a much stronger response than under the model based on conformational stability. 
This effect is due to less sites in the protein being involved in protein-protein interaction than for conformational stability, in addition to the lower energy engaged in the contacts.
As such, in light of empirical data available, the assumption that protein fitness is determined such as to avoid non-specific interaction is more likely than the assumption that protein fitness is determined by the deleterious effect of unfolded proteins. \\
% Protein-protein interactions

% DFE can also be used to fit to empirical data - Shall we present this ?
